---
title: "Reproducible research - Course Project 1"
author: "V.SH."
date: "March 13, 2019"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r libraries, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)
```
```{r load_data}
activity.raw <- read.csv("activity.csv")
str(activity.raw)
activity.raw %>% mutate(date=as.Date(date,"%Y-%m-%d")) -> activity
## sample
activity[c(1:3,17278:17280),]
str(activity)
## distribution of dates/intervals with missing values for steps
as.data.frame(table(activity[is.na(activity$steps)==1,2])) 
```
The activity data contains `r  17568/288` (17568/288) complete days, with 288=24*12 (24 hours, each contains twelve 5-minute intervals) measurements per day. The interval IDs are obviously in yymm format - read.csv reads them as integers and it's OK.   

The missing measurements are present for 8 dates and all step values for these eight days - two Mondays, one Wednesday, one Thursday, two Fridays, one saturday, and one Sunday (no Tuesdays)  - are missing. The  weekdays might be important, as many people have weekly routines.


## What is mean total number of steps taken per day?
```{r daily_aggregates}

activity %>% group_by(date) %>%summarise(totsteps=sum(steps)) -> daily
mm<-paste0("Mean is ", 
           format( mean(daily$totsteps,na.rm=T),nsmall=1)," and median is ",
           format( median(daily$totsteps,na.rm=T)))
hist(daily$totsteps, main="Total steps per day: distribution",xlab=mm,
     breaks=15)

```



## What is the average daily activity pattern?
```{r by_interval}
activity %>% filter(is.na(steps)==0) %>% group_by(interval) %>% 
    summarise(avgstep=mean(steps)) ->activity_by_int
with(activity_by_int, plot(interval,avgstep,type="l"))
## interval with max avgstep
maxstep<-activity_by_int[activity_by_int$avgstep==max(activity_by_int$avgstep),1]

```
Max average number of steps corresponds to `r maxstep`.


## Imputing missing values
As mentioned earlier, people often have weekly routines. Accordingly, for the days with missing step counts, I am using the averages for the same weekday-interval combination.
```{r imputting}
activity %>%
    mutate(wkday=wday(date,label=T)) %>% 
    group_by(interval,wkday) %>% 
    summarise(avgstep=mean(steps,na.rm=T)) ->activity_by_int_wkday

tmp<-merge(mutate(activity,wkday=wday(date,label=T)),activity_by_int_wkday)
# activityF - with filled-in missing values, and weekdays 
tmp %>% mutate(steps=ifelse(is.na(steps)==T,avgstep,steps)) %>%
    select(-avgstep) -> activityF

activityF %>% group_by(date) %>%summarise(totsteps=sum(steps)) -> dailyF
par(mfrow=c(1,2))

hist(dailyF$totsteps,main="Missing values filled in",xlab="total steps",
     ylim=c(0,20),breaks = 15)
hist(daily$totsteps,main="As reported",xlab="total steps",
     ylab="",ylim=c(0,20),breaks=15)

daily  %>% summarise(meanS=mean(totsteps,na.rm = T),
                            medS=median(totsteps,na.rm = T)) ->DD
dailyF %>% summarise(meanS=mean(totsteps),
                            medS=median(totsteps)) -> DDF

```
The histograms are similar, but not identical, and agrregate values are affected by imputting.  Before imputting, mean=`r format(DD$meanS, nsmall=1)` and 
median=`r format(DD$medS)`. After imputting, mean=`r format(DDF$meanS, nsmall=1)` and median=`r format(DDF$medS)`.  
After imputting, the aggregate values have increased, but not by much. 


## Are there differences in activity patterns between weekdays and weekends?
For this part of the assignment, I was required to use the dataset with the filled-in missing values and to create a new factor variable  with two levels – “weekday” and “weekend.”
```{r weekdays vs weekends}
activityF %>% mutate(isWkEnd=ifelse(wkday %in% c('Sat','Sun'),
                                    "weekend","weekday")) -> activityF2
activityF2 %>%
    group_by(interval,isWkEnd) %>% 
    summarise(avgstep=mean(steps)) ->activityF2_by_int_isWkEnd
# ggplot(activityF2_by_int_isWkEnd,aes(interval,avgstep)) ->g3
# g3+geom_line()+facet_grid(rows=vars(isWkEnd))
G1 <- ggplot(activityF2_by_int_isWkEnd, aes(interval,avgstep)) 
G1<- G1 + geom_line() + facet_grid(rows=vars(isWkEnd)) 
G1<- G1 + labs(y="average number of steps in an interval",
               title="weekdays vs. weekends")
G1 + theme(plot.title = element_text(hjust = 0.5))
```
  
The chart shows that, on average, the daily activity on weekends start and ends later than on weekdays.


